---
title: Wealth Inequality and Vibesssss
author: 7-UP
output: github_document
---

```{r load-packages, include = FALSE}
library(tidyverse)
library(viridis)
library(ggrepel)
```

```{r load-data, include = FALSE}
income_mean <- read_csv(file = "data/income_mean.csv")
home_owner <- read_csv(file = "data/home_owner.csv")
```


## Introduction

(1-2 paragraphs)
Brief introduction to the dataset.
You may repeat some of the information about the dataset provided in the introduction to the dataset on the TidyTuesday repository, paraphrasing on your own terms.
Imagine that your project is a standalone document and the grader has no prior knowledge of the dataset.

The dataset comes from the Urban Institute and US Census and focuses on racial 
wealth inequality. We chose this dataset because wealth inequality is a very 
important issue in America, and it would be valuable to see the relationship 
between race and wealth over time in America. There are around 5-6 data files 
focused on the relationship between race and income over time. There are also 
important files on the relationship between race and retirement, homeownership, 
and student debt.



## How has home-ownership changed over time, and how does this trend vary between different races?

### Introduction

(1-2 paragraphs)
Introduction to the question and what parts of the dataset are necessary to answer the question. Also discuss why you’re interested in this question.

We first wanted to investigate how home ownership has changed over time, and 
among different racial groups. We decided to use the `home_owner` dataset and 
use the `year`, `race`, and `home_owner_pct` variables. 

### Approach

(1-2 paragraphs)
Describe what types of plots you are going to make to address your question.
For each plot, provide a clear explanation as to why this plot (e.g. boxplot, barplot, histogram, etc.) is best for providing the information you are asking about. 
The two plots should be of different types, and at least one of the two plots needs to use either color mapping or facets.

### Analysis

(2-3 code blocks, 2 figures, text/code comments as needed)
In this section, provide the code that generates your plots. Use scale functions to provide nice axis labels and guides.
You are welcome to use theme functions to customize the appearance of your plot, but you are not required to do so.
All plots must be made with ggplot2. Do not use base R or lattice plotting functions.


```{r find-peaks, include = F}
lowest_year <- home_owner %>%
  group_by(year) %>%
  summarize(mean_pct_unweighted = mean(home_owner_pct)) %>%
  arrange(mean_pct_unweighted) %>%
  slice(1) %>%
  pull(year)
highest_year <- home_owner %>%
  group_by(year) %>%
  summarize(mean_pct_unweighted = mean(home_owner_pct)) %>%
  arrange(desc(mean_pct_unweighted)) %>%
  slice(1) %>%
  pull(year)

last_year <- home_owner %>%
  group_by(race) %>%
  filter(year == max(year))
```

#### Plot 1.a
```{r first-look, echo = F}
home_owner %>%
  ggplot(
    aes(x = year, y = home_owner_pct, color = race, shape = race)
  ) +
  geom_line(size = 1) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_vline(aes(xintercept = lowest_year), color = "grey") +
  geom_vline(aes(xintercept = highest_year), color = "grey") +
  annotate(geom = "text",
    x = lowest_year - 5, 
    y = 0.56,
    label = paste0("Year with lowest \nhome ownership (", lowest_year, ")"),
    color = "black"
    ) +
  annotate(
    geom = "text",
    x = highest_year + 5, 
    y = 0.56, 
    label = paste0("Year with highest \nhome ownership (", highest_year, ")"), 
    color = "black"
    ) +
  geom_text_repel(data = last_year,
                  aes(label = race),
                  direction = "y",
                  nudge_x = 2,
                  nudge_y = 0.02,
                  size = 5,
                  segment.linetype = "dotted") +
  scale_color_manual(values = c("#859a6a", "#ad6e72", "#5d7a96")) +
  scale_y_continuous(labels = scales::percent_format(scale = 100)) +
  labs(
    x = "Year",
    y = "Percent owning homes",
    color = "Race",
    shape = "Race",
    title = "Home ownership over time",
    subtitle = "by race",
    caption = "7 Up: Blossom Mojekwu, Kartik Chamarti, Margaret Reed, Phillip Harmadi"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Discussion

(1-3 paragraphs)
In the Discussion section, interpret the results of your analysis. Identify any trends revealed (or not revealed) by the plots.
Speculate about why the data looks the way it does.

## What does household income level look like for each racial group over time, and how do household income disparities compare among different racial groups?

### Introduction

(1-2 paragraphs)
Introduction to the question and what parts of the dataset are necessary to answer the question. Also discuss why you’re interested in this question.

Most income inequality research focuses on the income disparity between 
different races, for example, how Asian Americans and White Americans typically 
have higher income as compared to African Americans and Hispanic/Latino 
Americans. There is significantly less research focusing on the income 
disparities within each of the races. We are trying to figure out the degree of 
income inequality for each of the races within their own community.
In this question, we will be using `income_mean.csv`. This data set is 
sufficient since it provides us with the income information for each household 
quintile, year, and race. We will be using the following variables: `year`, 
`race`, `income_quintile`, and `income_dollars` which will be explained further 
in the "Approach" section.


### Approach

(1-2 paragraphs)
Describe what types of plots you are going to make to address your question.
For each plot, provide a clear explanation as to why this plot (e.g. boxplot, barplot, histogram, etc.) is best for providing the information you are asking about. 
The two plots should be of different types, and at least one of the two plots needs to use either color mapping or facets.

In order to investigate how average income distribution by race has changed over 
time we made a line plot faceted by race with year on the x axis and average 
income amount on the y axis. We decided to do so, because we cared less about 
looking into specific years and more about overall trends and comparisons within 
each race. To clean the data for this visualization, we filtered out race 
categories that included combinations and the “all” category. This left us with 
4 levels: white, black, asian, hispanic. We wanted to highlight the top earners 
in each racial category- so we highlighted them in red. Something interesting to 
note is that data collection for asian observations only started around 2000. It 
is hard to easily compare differences between races, so we decided to zoom into 
2019 and make a bar chart.

**Average income for each income quintile by race.** A ***bar graph*** is deemed 
appropriate for this plot since we are trying to compare how much the top income 
quintile earns more than the bottom income quintile for each racial group and 
understand which race has the highest gap. A bar graph is ideal since we can 
group or stack them together and color them based on the income quintile 
variable, making the visualization easier to be understood by readers.



### Analysis

(2-3 code blocks, 2 figures, text/code comments as needed)
In this section, provide the code that generates your plots. Use scale functions to provide nice axis labels and guides.
You are welcome to use theme functions to customize the appearance of your plot, but you are not required to do so.
All plots must be made with ggplot2. Do not use base R or lattice plotting functions.


```{r data-wrangle-2, include = F}
income_mean <- income_mean %>%
  filter(!str_detect(race, "Combination"),
         !str_detect(race, "Not Hispanic"),
         !str_detect(race, "All"),
         dollar_type == "2019 Dollars",
         income_quintile != "Top 5%") %>%
  group_by(year, race, income_quintile) %>%
  summarize(income_dollars =  mean(income_dollars), 
            .groups = "drop") %>%
  mutate(race = str_remove(race, ',?\\s(.*)'),
         income_quintile = ordered(income_quintile, c("Lowest", "Second", 
                                                      "Middle", "Fourth", 
                                                      "Highest")),
      income_percentile = case_when(
        income_quintile == "Top 5%"   ~ "1st - 5th",
        income_quintile == "Highest"  ~ "Top 20%",
        income_quintile == "Fourth"   ~ "21-40 %",
        income_quintile == "Middle"   ~ "41-60 %",
        income_quintile == "Second"   ~ "61-80 %",
        income_quintile == "Lowest"   ~ "Bottom 20%"),
        income_percentile = ordered(income_percentile, 
                                  c("Bottom 20%", "61-80 %", "41-60 %", 
                                    "21-40 %", "Top 20%")))
```

```{r perc-plot, include = F}
highest_perc_table <- income_mean %>%
  filter(year == 2019) %>%
  group_by(race) %>%
  transmute(income_quintile, percent = income_dollars/sum(income_dollars)) %>%
  filter(income_quintile == "Highest")
```



#### Plot 2.a

```{r plot-overtime, echo = F}
income_mean %>%
  ggplot(
    aes(x = year, y = income_dollars, color = income_percentile)
  ) +
  geom_line() +
  facet_wrap(.~race, nrow = 1) +
  scale_y_continuous(labels = scales::label_number(big.mark = ",", 
                                                   scale = 0.001, 
                                                   suffix = "k",
                                                   prefix = "$")) +
  scale_color_manual(values = c("grey40", "grey50", "grey60", "grey70", 
                                "firebrick2")) +
  labs(
    title = "Income Gap: How does income distribution compare for each race?",
    subtitle = "By income quintile",
    x = "Year", 
    y = "Mean household income", 
    color = "Income quintile",
    caption = "7 Up: Blossom Mojekwu, Kartik Chamarti, Margaret Reed, Phillip Harmadi"
    ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

#### Plot 2.b

```{r plot-bar, echo = F}
income_mean %>%
  filter(year == 2019) %>%
  ggplot(
    aes(x = income_dollars,
        y = reorder(race, desc(race)))
    ) + 
  geom_bar(
    position = "fill", 
    stat = "identity", 
    aes(fill = reorder(income_percentile, desc(income_percentile)))
    ) +
  geom_text(
    data = highest_perc_table, 
    aes(
      y = race, label = paste0(round(percent, 2)*100, "%")),
    x = 0.8, 
    color = "white") +
  scale_fill_brewer(type = "seq", palette = "Blues", direction = -1) +
  labs(
    title = "Income Gap: How much do top households earn compared to bottom households \nfor each race?",
    subtitle = "In 2019",
    x = "Percent of total mean income", 
    y = "Race", 
    fill = "Income Quintile",
    caption = "7 Up: Blossom Mojekwu, Kartik Chamarti, Margaret Reed, Phillip Harmadi"
    ) +
  scale_colour_viridis_d() +
  scale_x_continuous(labels = scales::percent_format(scale = 100)) +
  theme_minimal(base_size = 10) +
  theme(text = element_text(size = 9))
```


### Discussion

(1-3 paragraphs)
In the Discussion section, interpret the results of your analysis. Identify any trends revealed (or not revealed) by the plots.
Speculate about why the data looks the way it does.


From the first plot, we found that those who are asian tended to have the 
highest average income, and their top 20% made far more than that of the rest of 
the racial groups. Overtime, income disparities in all the racial categories 
increased when considering the top 20% of income earners. This increase in 
disparity seemed to be most noticeable for the white observations. We suspect 
that income is the highest among Asians due to their high educational attainment 
on average, and lowest in Blacks due to the systemic disadvantages they have 
experienced throughout U.S. history.
From the second plot, we can roughly surmise that income inequality is the 
highest within Black households, followed by Asian, White, then Hispanic 
households. Our speculation is that income inequality is highest among the 
Black category on average,  because of lack of high-paying jobs and 
opportunities for Black individuals living in the rural Southern States vs. 
those living in Northeastern or Midwestern urban areas. For Asians, we speculate 
that income inequality comes from different migration patterns and sources. For 
example, Indian and Chinese Americans have a high proportion of professional 
college-educated citizens, while a lot of Southeast Asian Americans are doing 
blue-collared jobs and many came to the United States as refugees from less 
wealthy backgrounds.


## Presentation

Our presentation can be found [here](presentation/presentation.html).

## Data 

Include a citation for your data here. 
See http://libraryguides.vu.edu.au/c.php?g=386501&p=4347840 for guidance on proper citation for datasets. 
If you got your data off the web, make sure to note the retrieval date.

## References

List any references here. You should, at a minimum, list your data source.

Our data comes from the [Urban Institute](https://apps.urban.org/features/wealth-inequality-charts/) 
and the [US Census](https://www.census.gov/data/tables/time-series/demo/income-poverty/historical-income-households.html). 

The US Census provides [Historical Income Tables](https://www.census.gov/data/tables/time-series/demo/income-poverty/historical-income-households.html), of which we have joined several to compare wealth and income over time by 
race.

