---
title: "Exploring income equality"
author: "Seven-Up"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
```

## Dataset

```{r load-data, message = FALSE, warning = FALSE}
home_owner <- read_csv("../data/home_owner.csv")
income_distribution <- read_csv("../data/income_distribution.csv")
```

The dataset comes from the Urban Institute and US Census and focuses on racial
wealth inequality. We chose this dataset because wealth inequality is a very
important issue in America, and it would be valuable to see the relationship
between race and wealth over time in America. There are around 5-6 data files
focused on the relationship between race and income over time. There are also
important files on the relationship between race and retirement, homeownership, 
and student debt. 

We anticipate primarily using the `home_owner` dataset and the 
`income_distribution` dataset. The `home_owner` table has `r dim(home_owner)[1]` 
rows and `r dim(home_owner)[2]` columns. The `income_distribution` table has `r dim(income_distribution)[1]` rows and `r dim(income_distribution)[2]` columns. 

## Questions

**Question 1:** How has homeownership changed over time, and how does this trend 
vary between different races?

We hypothesis that homeownership has decreased overtime and possibly that the 
differences in home ownership among races have decreased over time. We would 
like to further explore these hypotheses in our visualizations and analysis. We 
anticipate using some sort of time series plot.

**Question 2:** How is wealth and income distributed within different racial 
groups, and how has this changed over time?

Our hypothesis is that within racial groups, the wealth disparities has actually 
increased and that overall between racial groups the disparities are 
significant. 


## Analysis plan

A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).
