---
title: "Income Equality in the United States"
author: "7 Up: Blossom Mojekwu, Kartik Chamarti, Margaret Reed, Phillip Harmadi"
output: github_document
---

```{r load-packages, message = FALSE, warning = FALSE}
library(tidyverse)
```

## Dataset

```{r load-data, message = FALSE, warning = FALSE}
home_owner <- read_csv("../data/home_owner.csv")
income_mean <- read_csv("../data/income_mean.csv")
```

The dataset comes from the Urban Institute and US Census and focuses on racial
wealth inequality. We chose this dataset because wealth inequality is a very
important issue in America, and it would be valuable to see the relationship
between race and wealth over time in America. There are around 5-6 data files
focused on the relationship between race and income over time. There are also
important files on the relationship between race and retirement, homeownership, 
and student debt. 

We anticipate primarily using the `home_owner` dataset and the 
`income_mean` dataset. The `home_owner` table has `r dim(home_owner)[1]` 
rows and `r dim(home_owner)[2]` columns. The `income_mean` table has 
`r dim(income_mean)[1]` rows and `r dim(income_mean)[2]` columns. 

## Questions

**Question 1:** How has homeownership changed over time, and how does this trend 
vary between different races?

- We hypothesis that homeownership has decreased overtime and possibly that the 
differences in home ownership among races have decreased over time. We would 
like to further explore these hypotheses in our visualizations and analysis. We 
anticipate using some sort of time series plot.

**Question 2:** How does household income level growth look like for each racial
group over time, and how does the household income disparities look like among 
each racial group?

- Our hypothesis is that within racial groups, the income disparities 
has actually increased and that overall between racial groups the disparities 
are significant. We also hypothesize that household income inequalities is the 
highest within Asian-Americans. Our reasoning is that Asian-Americans came from 
different migration purposes. A significantly higher proportion of East-Asian 
and Indian-Americans came as professional and skilled white-collared workers 
while a higher proportion of Southeast-Asian Americans and Pacific Islander 
Americans came as blue-collared workers or war refugees.

## Analysis plan

A plan for answering each of the questions including the variables involved, 
variables to be created (if any), external data to be merged in (if any).

**Question 1** will visualize how home owner ship changes over time, and how it
differs among races. 

- We plan to use all of the variables from the `home_owner` dataset: `race`, 
`year`, and `home_owner_pct`.

- We don't anticipate needing additional data, however it is possible we might 
include data such as home vs rent prices over time.

- In regards to how this research question will be displayed, we could create
a line plot across 3 different races so that the plot contains 3 different lines
of different races to compare for context. We are also considering including 
markers for when significant moments for the housing market occured (e.g. the 
stock market crash).

**Question 2** will visualize the income figures for each of household income 
quintile - how their figure changes over time for each racial group.

- Variables involved will be `race`, `year`, `income_percentile` (mutated from 
`income_quintile`), and household income (in 2019 Dollars for the first figure 
and in ratio to the lowest quintile income per year and per race for the 
second figure).

- There will be no external data to be merged in, but in the process, we will 
expect to perform some amount of data wrangling, e.g., `mutate`, `filter`, 
`select`, and `pivot`.

- We plan to use boxplots or bar charts to show how the distribution of income 
differs among different races. To show change over time we would include a 
separate visualization. The interpretation for this plot would be connected to 
that from question 1.

